---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# madisonamg

<!-- badges: start -->
<!-- badges: end -->

The goal of `madisonamg` is to provide a simplified pathway for loading AMG data, checking for any issues, finding peaks of the response, analyzing the data, and visualizing it. `madisonamg` (Madison AMG) is named in honor of Madison, a golden retriever who passed away from myasthenia gravis, an autoimmune neuromuscular disease.

## Installation
You can install `madisonamg` from GitHub with [remotes](https://remotes.r-lib.org):

```{r remotes, eval = FALSE}
# install.packages("remotes")
remotes::install_github("jessesadler/madisonamg")
```

Please open an [issue](https://github.com/jessesadler/madisonamg/issues) if you have any questions, comments, or requests.

## Example analysis

This document shows the basic capabilities of the `madisonamg` package and how an analysis of a trace might be performed.

```{r load-package}
library(madisonamg)
```

### The data

The data is expected to be imported as a wave file using `tuneR::readWave("path/file.wav", toWaveMC = TRUE)`. Make sure to set `toWaveMC = TRUE`, which imports multiple tracks, since `madisonamg` functions expect readings for the amplitude of both the stimulus and the response. `madisonamg` provides an example wave file imported in this fashion: `ex_waveMC`. Let's load it.

```{r load-data}
wav <- ex_waveMC
```

To work with the data we need to get it into the form of a data frame with `waveMC_to_tbl()`. This will make it much easier to use the normal data analysis pipeline of R and the [tidyverse](https://www.tidyverse.org) on the AMG data. `waveMC_to_tbl()` also creates a data frame, or [tibble](https://tibble.tidyverse.org/), that conforms to the structure expected by the functions in `madisonamg`. We can refer to this as a trace tibble. `waveMC_to_tbl()` creates a tibble with four columns: sample, secs, stimulus, and response. The function automatically distinguishes the stimulus recording channel from the response recording channel. The frequency (`freq` argument) of the recording is used to convert samples into seconds (secs). The secs variable is there for convenience. It makes it easier to see how long the recording is, where events are in the recording, or how far apart they are.

Let's make a trace tibble with `waveMC_to_tbl()`. This is equivalent to the example data `ex_trace_tbl` which comes with `madisonamg`.

```{r create-tbl}
trace_tbl <- waveMC_to_tbl(wav, freq = 10000)

trace_tbl
```

Most functions from `madisonamg` expect a trace tibble as an input. A **trace tibble** is a data frame or tibble with numeric columns named "sample", "stimulus", and "response" plus any other metadata columns such as secs. If any of these three columns are not present or if they have different names, the functions in `madisonamg` will not work.

### Examing the data

The easiest way to inspect the data is to start by visualizing it. We can separately visualize the recording of the response and stimulus to make sure there are no oddities. Setting `filter_response = FALSE` visualizes the entire recording.

```{r full-plots}
viz_response(trace_tbl, filter_response = FALSE, title = "Full response")

viz_stimulus(trace_tbl, filter_stimulus = FALSE, title = "Full stimulus")
```

Both of the plots look good. They both have ten clearly distinguished peaks. The response plot shows the points in red where the stimuli were performed and these seem to correspond to the peaks. We can get a closer look at the shape of the response by using the same function but setting `filter_response` to `TRUE`, which is the default. This filters out the response before the first stimulus and after the last stimulus.

```{r filtered-reponse}
viz_response(trace_tbl, title = "Response")
```

Note that you can change the aesthetic features of the plot by adding `ggplot2` commands. For instance, we can focus more on the response by getting rid of extra elements through `theme_void()`. I also add a bit more data before to each side of the plot with `buffer`. Make sure to load `ggplot2`.

```{r ggplot}
library(ggplot2)
viz_response(trace_tbl,
             title = "Response",
             show_stimulus = FALSE,
             buffer = 1000) + 
  theme_void()
```

Another very useful way to explore the data through visualization is with `viz_response_interactive()`, which creates an interactive plot of the response data, enabling the user to zoom in on the data.

### Finding the stimuli and peaks

The next step is to find when the stimuli performed and calculate the starting and ending point for the corresponding response peaks. Finding the stimuli is relatively simple because when the stimuli is performed the amplitude of the stimulus rises immediately from its baseline. This is how `find_stimuli()` works, using the `stimulus_diff` argument to provide a minimum amplitude difference between samples.

```{r find-stimuli}
find_stimuli(trace_tbl)
```

In addition to finding the number of stimuli and the samples where they were performed, `find_stimuli(trace_tbl)` shows the time gap between stimuli and the calculated hertz. In this case, we can see that the stimuli were performed once per second, or 1hz.

Finding the peaks of the response is a bit more difficult and not nearly as clear cut. Looking at the response column in the output from `find_stimuli(trace_tbl)`, you can see that the response amplitude differs between -128 and 448. There is no completely consistent baseline. It can, therefore, be difficult to determine when the amplitude of the response begins to rise as a result of the stimulus. `madisonamg` provides three functions to help find peaks: `find_peaks_response()`, `find_peaks_stimulus()`, and `find_peaks_manual()`. These functions should most likely be used in this order.

`find_peaks_response()` is the most robust method and works well when the peaks are relatively uniform. It works by finding response amplitudes above a specified `min_amp` and then extends the region down to all samples above the given `baseline`. The defaults all work pretty well with the `trace_tbl` data, but they may have to be altered for different data. In particular, the `baseline` argument can be manipulated to have unique baselines for each peak if the data is a bit messier.

It should also be noted that instead of returning a tibble, the `find_peaks_()` functions return a list of numeric vectors of the samples where the peaks occur. This data type is used throughout the `madisonamg` package wherever there is a `peaks` argument.

```{r find-peaks-response}
peaks <- find_peaks_response(trace_tbl)
```

`find_peaks_stimulus()` is a simpler function. It finds the onset of the stimuli in a fashion similar to `find_stimuli()` and starts the peak a set number of milliseconds after each stimulus as defined by the `delay` argument. The `baseline` argument can either be defined by the user or decided separately for each peak based on the value of the response after the given delay. This latter method is the default.

Finally, `find_peaks_manual()` is a helper function to manually input the start and end of peaks. This is useful for particularly odd peaks. The starting and ending points can be determined through visualizing the peaks. `viz_response_interactive()` is particularly useful here.

Because `find_peaks_response()` works well here, we will use that to define `peaks`.

### Checking the peaks

The `find_peaks_()` functions will define the beginning and end of peaks, but that does not mean that they will correctly find them. You can perform a number of checks to ensure that the peaks were correctly identified. The simplest check is to ensure that the number of peaks identified is the same as the number of stimuli identified. We can then move to see what the delay is between the onset of the stimuli and the beginning of the peaks.

```{r peaks-delay}
length(peaks) == nrow(find_stimuli(trace_tbl))
peaks_delay(trace_tbl, peaks)
```

`peaks_delay()` shows that the fifth peak is a bit of an outlier with a delay of only 3.1 milliseconds. This is somewhat unsurprising as the fifth stimulus occurred when the response amplitude was at 448.

We can also make some other checks on the peaks data to see what we are dealing with. Here, we are looking for consistency. However, it is important to note that the amplitude values are discontinuous.

```{r peaks-checks}
peaks_min_amp(trace_tbl, peaks)
peaks_max_amp(trace_tbl, peaks)
peaks_first_amp(trace_tbl, peaks)
peaks_last_amp(trace_tbl, peaks)
```

Looking at the results for `peaks_last_amp(trace_tbl, peaks)`, we can see that the amplitudes often do not get down to the baseline. We can fix this by augmenting the `lengthen` argument in `find_peaks_response()`.

```{r find-peaks-2}
peaks2 <- find_peaks_response(trace_tbl, lengthen = 200)
```

Rechecking the peaks with `peaks2`, we can see that the end of the peaks is now correct. However, the fifth peak now starts before the stimulus, which is clearly wrong.

```{r peaks-checks-2}
peaks_last_amp(trace_tbl, peaks2)
peaks_delay(trace_tbl, peaks2)
```

### Visual checks on the peaks

The other primary manner of checking the peaks is to do so visually. The easiest first step to double check that the peaks broadly correspond to the stimuli is with `viz_highlight_peaks()`.

```{r highlight-peaks}
viz_highlight_peaks(trace_tbl, peaks2, title = "Highlight peaks")
```

To get into a more in depth visual analysis of the peaks, we can use `viz_peak()` to visualize individual peaks and provide information about them. Let's look at the fifth peak.

```{r viz-peak5}
viz_peak(trace_tbl, peaks2, peak_nr = 5)
```

You could also get an even closer view with `viz_response_interactive(trace_tbl, peaks)` and zooming in on the fifth peak. Using this method, it is possible to see that the baseline of the fifth peak is 448. At this point, you could manually create a peak with `find_peaks_manual()` and inputting the samples where the amplitude goes above 448 and then back below 448. Or you can create separate baselines in `find_peaks_response()`, which we will do here.

```{r find-peaks-3}
peaks3 <- find_peaks_response(trace_tbl, lengthen = 200,
                              baseline = c(rep(256, 4), 448, rep(256, 5)))
```

We can then check this change.

```{r}
peaks_first_amp(trace_tbl, peaks3)
peaks_last_amp(trace_tbl, peaks3)
peaks_delay(trace_tbl, peaks3)
```

The numbers are much more consistent now. Let's see how it looks visually.

```{r viz-peak5-2}
viz_peak(trace_tbl, peaks3, peak_nr = 5)
```

### Analyzing the data

Finally, we can analyze the peak data by creating an AMG report with `create_report()`. This shows the peak amplitude of each peak and the area under the curve along with percentage decrease in relation to the first peak. The area is calculated using the minimum value of each peak as the baseline, but you can also provide your own through the `baseline` argument.

```{r report}
create_report(trace_tbl, peaks3)
```
